---
title: "Visualize Data"
output:
  html_document:
    df_print: paged
---

<!-- This file was adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup}
library(tidyverse)
library(here)
```


```{r}
wq <- read.csv(here::here('data', 'monthly_wq.csv')) 

wq

dplyr::glimpse(wq)
```

## Quiz

What relationship do you expect to see between temperature (temp) and dissolved oxygen (do_mgl)?

## Your Turn 1

Run the code on the slide to make a graph. Pay strict attention to spelling, capitalization, and parentheses!

```{r}
ggplot(data = wq) +
  geom_point(mapping = aes(x = temp, y = do_mgl))
```

## Your Turn 2

Add `color`, `size`, `alpha`, and `shape` aesthetics to your graph. Experiment.  

```{r}
ggplot(data = wq) + 
  geom_point(mapping = aes(x = temp, y = do_mgl, color = 'blue'), size = 3)
```

## Your Turn 3

Replace this scatterplot with one that draws boxplots. Use the cheatsheet. Try your best guess.

```{r}
ggplot(data = wq) +
  geom_point(mapping = aes(x = coast, y = temp))

ggplot(data = wq) +
  geom_boxplot(mapping = aes(x = coast, y = temp))

```

## Your Turn 4

Predict what this code will do. Then run it.

```{r}
ggplot(wq) +
  geom_point(aes(temp, do_mgl)) +
  geom_smooth(aes(temp, do_mgl))
```

## Your Turn 5

What does `getwd()` return?

```{r}
getwd()
```

## Your Turn 6

Save the last plot and then locate it in the files pane.

```{r}

```

***

# Take aways

You can use this code template to make thousands of graphs with **ggplot2**.

```{r, eval = FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```
