---
title: "Visualize Data"
output:
  html_document:
    df_print: paged
---

<!-- This file by Shannon Dunnigan was adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup}
library(tidyverse)
library(here)
```


```{r loaddata}
wq <- read.csv(here::here('data', 'monthly_wq.csv')) 

```

## Quiz

What relationship do you expect to see between temperature (temp) and dissolved oxygen (do_mgl)?

## Your Turn 1

Run the code on the slide to make a graph. Pay strict attention to spelling, capitalization, and parentheses!

```{r}


```

## Your Turn 2

Add `color`, `size`, `alpha`, and `shape` aesthetics to your graph. Experiment.  

```{r}
ggplot(data = wq) + 
  geom_point(mapping = aes(x = temp, y = do_mgl))

```

## Your Turn 3

Using the `wq` data, plot temperature (`temp`) by `month` as a scatterplot and facet by station (`station_code`), `state`, and `coast`. 
Play around with `facet_grid()` and `facet_wrap()`. 

*What kinds of things do you notice?*

```{r}
ggplot(data = wq) +
  geom_point(mapping = aes(x = month, y = temp))

```

## Your Turn 4

Replace this scatterplot with one that draws boxplots. Use the cheatsheet. Try your best guess.

```{r}
ggplot(data = wq) +
  geom_point(mapping = aes(x = coast, y = temp)) 

```

## Your Turn 5

Load the `diamonds` dataset and examine it. *ahem* Remember the `glimpse()` function? This data set contains prices and other attributes of diamonds and is provided by the `ggplot2` package.

Create a bar chart of cut colored by cut. Use the cheatsheet. 

```{r}
diamonds

```

## Your Turn 6

Determine what the default stat is for each of the following geoms:

`geom_line`
`geom_density`
`geom_smooth`

```{r}


```

## Your Turn 7

Predict what this code will do. Then run it.

```{r}
ggplot(wq) +
  geom_point(aes(temp, do_mgl)) +
  geom_smooth(aes(temp, do_mgl)) 

```

## Your Turn 10 

What does `getwd()` return?

```{r}
getwd()
```

## Your Turn 11

Save the last plot and then locate it in the files pane.

```{r}

```

## MINI CHALLENGE

Using the diamonds data make a scatterplot of carat by price and color by clarity. 
Customize the themes and colors of the plot to make it the *ugliest* plot that you can. 

``` {r ugly}
ggplot(data = diamonds) +
  geom_point(aes(x = carat, y = price, color = clarity))
```

***

# Take aways

You can use this code template to make thousands of graphs with **ggplot2**.

```{r, eval = FALSE}
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>),
                  stat = <STAT>,
                  position = <POSITION>) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION> +
  <SCALE_FUNCTION> +
  <THEME_FUNCTION>
```
